---
title: "R-Ladies RTP ggplot2 Workshop"
author: "Sheila Saia"
date: "March 6, 2019"
output:
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome to R-Ladies RTP!

## Outline

* Why ggplot2?
* Conceptual ggplot2
* Basic ggplot2 syntax
* Common pitfalls
* The power of layers
* The power of groups
* The power of facets
* Scales and legends
* Themes and refinements
* Extensions

## Materials

All source code and slides at [TBD]

## Datasets

* Datasets for today are here:
https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1

* You can download them from R:

    `download.file(
      'https://www.dropbox.com/s/zhmn02ti0ggxdj7/
      rladies_ggplot2_datasets.rda?dl=1',
      'rladies_ggplot2_datasets.rda')`
    
    `attach('rladies_ggplot2_datasets.rda')`
  
## Dataset Descriptions

* Transit ride data
    + rides_df: one row per ride
    + daily_df: daily summary of rides
    + hourly_df: hourly summary of rides
  
* Durham registered voter data
    + durham_voters_df: one row per voter

# Why ggplot2?

## Exploratory Power

* explore data very quickly, and don't screw up
* graphics "pit of success"
* refine plots to be publication quality
* see discussions of ggplot2 vs base R
    + http://nssdeviations.com/episode-12-the-new-bayesian-vs-frequentist

## Conceptually Cohesive

* based on The Grammar of Graphics by L. Wilkinson and others
* principles and patterns
* very complex plots from simple building blocks
    + https://vita.had.co.nz/papers/layered-grammar.pdf

## Dominance

* the most prominent static graphics package in R
* help is easy to find
* developers are contibuting extensions

# Conceptual ggplot2

## The basics

* map variables to aethestics
* add "geoms" for visual representation layers
* scales can be independently managed
* legends are automatically created
* statistics are sometimes calculated by geoms...

## Learn more

* See the ggplot2 book by H. Wickham
* See the R for Data Science book

# Basic ggplot2 syntax

## The required elements

* DATA
* MAPPING
* GEOM

## A required element aside: geom versus stat

* A geom is *required*. A stat (i.e., statistical tranformation) is *not required*.
* We won't cover stat objects today...just letting you know they exist.

For example:
  `geom_bar()` uses `stat_count` as it's default stat.

## Install the tidyverse

Run this only if it's your first time using `tidyverse` on your computer.

```{r, message=FALSE, eval=FALSE}
install.packages("tidyverse")
```

## Load the tidyverse

```{r, message=FALSE}
library("tidyverse")
```

## Our first hands-on example ...

```{r, message=FALSE, eval=FALSE} 
download.file(
'https://www.dropbox.com/s/zhmn02ti0ggxdj7/rladies_ggplot2_datasets.rda?dl=1',
'rladies_ggplot2_datasets.rda')
        
attach('rladies_ggplot2_datasets.rda')

# or download from dropbox link and then run
# load("<your path to the .rda file>")
```

```{r, echo=FALSE}
load("/Users/ssaia/Downloads/rladies_ggplot2_datasets.rda") # for example
```

## Basic plot

```{r}
ggplot(data = daily_df) +
geom_point(mapping = aes(x = ride_date, y = n_rides))
```

## Mappings can be general

```{r}
ggplot(data = daily_df, mapping = aes(x = ride_date, y = n_rides)) +
geom_point()
```

## Parameters can be unnamed

```{r}
ggplot(daily_df, aes(x = ride_date, y = n_rides)) +
geom_point()
```

## Data can be passed in

```{r, fig.height = 4}
daily_df %>%
ggplot(aes(x = ride_date, y = n_rides)) +
geom_point()
# How else could we write this (using data = )?
```

## Data can be passed in

```{r, fig.height = 4}
hourly_df %>%
ggplot(aes(x = hour, y = n_rides)) +
geom_point()
```

## Mapping is more than x and y

```{r, fig.height = 4}
# Color by days of the week.
daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
geom_point()
```

## Mapping is more than x and y

```{r, fig.height = 4}
# Size by number of riders.
daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, size = n_riders)) +
geom_point()
```

## Variable creation on the fly...

```{r, fig.height = 4}
# Color by weekend or not.
daily_df %>%
ggplot(aes(x = ride_date, y = n_rides, 
           color = day_of_week %in% c('Sat', 'Sun'))) +
geom_point()
```

## ... or passed in

```{r, fig.height = 3}
# Color by day type.
daily_df %>%
  mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%
ggplot(aes(x = ride_date, y = n_rides, color = day_type)) +
geom_point()
```

# Common early pitfalls

## Mappings that aren't

```{r, fig.height = 4}
daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides, color = 'blue'))
```

## Mappings that aren't

```{r, fig.height = 4}
daily_df %>%
ggplot() +
geom_point(aes(x = ride_date, y = n_rides), color = 'blue')
```

## + and %>%

`daily_df %>%`    
    `mutate(day_type = if_else(day_of_week %in% c('Sat', 'Sun'),
                            'Weekend',
                            'Weekday')) %>%`    

`ggplot(aes(x = ride_date, y = n_rides, color = day_type)) %>%`   

`geom_point()`

## Basic plot exercise

Plot the number of unique routes per day over time, colored by day of week. (n_unique_routes)

## Basic plot exercise

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_unique_routes, color = day_of_week)) +
  geom_point() 
```

# The power of layers 

## Basic plot

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() 
```

## Two layers!

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() +
  geom_line()
```

## Iterate on layers

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() + 
  geom_smooth(span = .1) # try changing span
```

## The power of groups

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() + 
  geom_line()
```

## Now we've got it

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(span = .2, se = FALSE)
```

## Control data by layer

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point(data = filter(daily_df,
                           !(day_of_week %in% c('Sat', 'Sun')) 
                           & n_rides < 200),
             size = 5, color = 'gray') +
  geom_point()
```

## Control data by layer {.smaller}

```{r, fig.height = 3}
low_weekdays_df <- daily_df %>%
  filter(!(day_of_week %in% c('Sat', 'Sun')) & n_rides < 100)

daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, 
             color = day_of_week, label = ride_date)) +
  geom_point(data = low_weekdays_df, size = 5, color = 'gray') +
  geom_text(data = low_weekdays_df, aes(y = n_rides + 15), 
            size = 2, color = 'black') +
  geom_point() 
# Note how we define the data here.
```

## Other geoms to layer

There are a number of other geoms besides `geom_point()`, `geom_line()`, `geom_smooth()`, and `geom_text()`.

More info: https://ggplot2.tidyverse.org/reference/

# The power of facets

## facet_wrap

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides)) +
  geom_point() + 
  geom_line() +
  facet_wrap( ~ day_of_week)
```

## facet_grid {.smaller}

```{r, fig.height = 3}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_bar(stat = 'identity') +
  facet_grid(race_code ~ gender_code)
```

## Free scales {.smaller}

```{r, fig.height = 3}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_bar(stat = 'identity') +
  facet_grid(race_code ~ gender_code, scales = 'free_y')
```

## Facets + layers {.smaller}

```{r, fig.height = 4, echo = FALSE}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_point() +
  geom_line(aes(x = age_cat)) +
  geom_line(aes(x = age_cat, y = n_rep), color = 'red') +
  geom_point(aes(y = n_rep), color = 'red') +
  facet_grid(race_code ~ gender_code, scales = 'free_y') +
  expand_limits(y = 0) 
```

## Facets + layers {.smaller}

Note: better to use gather

```{r, fig.height = 3, eval = FALSE}
durham_voters_df %>%
  group_by(race_code, gender_code, age) %>%
  summarize(n_voters = n(),
            n_rep = sum(party == 'REP')) %>%
  filter(gender_code %in% c('F','M') & 
           race_code %in% c('W', 'B', 'A') &
           age != 'Age < 18 Or Invalid Birth Date') %>%
  mutate(age_cat = as.numeric(as.factor(age))) %>%
  ggplot(aes(x = age, y = n_voters)) +
  geom_point() +
  geom_line(aes(x = age_cat)) +
  geom_line(aes(x = age_cat, y = n_rep), color = 'red') +
  geom_point(aes(y = n_rep), color = 'red') +
  facet_grid(race_code ~ gender_code, scales = 'free_y') +
  expand_limits(y = 0) 
```


# Everyday conveniences

## Labels and titles

```{r, fig.height = 3, message=FALSE}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_smooth(span = .2, se = FALSE) +
  xlab('') +
  ylab('# of Transit Rides') +
  ggtitle('Transit Rides over time by Day of Week') +
  scale_color_discrete('Day of Week')
```

# Scales and legends

## Scale transformation

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_reverse()
```

## Scale transformation

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_sqrt()
```

## Scale details

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  scale_y_continuous(breaks = c(0, 200, 500))
```

# Themes and refinements

## Overall Themes

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_bw()
```

## Overall Themes

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme_dark()
```

## Changing Details

```{r, fig.height = 4}
daily_df %>%
  ggplot(aes(x = ride_date, y = n_rides, color = day_of_week)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90))
```

## Themes Vignette

To really master themes:

http://docs.ggplot2.org/dev/vignettes/themes.html

# Documentation

## Online Docs

Better than the regular documentation:

http://docs.ggplot2.org/current/

Don't forget the `ggplot2` cheatsheet!

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

# Extensions

## Contributed extensions

Check it out!

http://www.ggplot2-exts.org/

# Any other questions about ggplot2?

## Next Meetup

Wednesday, March 20th, 6pm

* Location:
    +TBD, near South Durham
* Topic:
    +`#tidytuesday` Work Group

Topic/format suggestions?

## Thanks

Thanks to Elaine McVey for sharing her slides from a previous R-Ladies RTP meetup!