---
title: "Zillow API"
author: "Stephanie Zimmer"
date: "November 26, 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Overview

- Zillow API Start-up 
- ZillowR Functions
- Zillow API Output for a few functions


## Zillow API Start-up

- Get an API Key (https://www.zillow.com/howto/api/APIOverview.htm)
- Install ZillowR and XML packages, tidyverse is also used but not completely necessary

```{r, echo=TRUE}
library(ZillowR)
library(XML)
library(tidyverse)
```

## ZillowR Functions
Function | Brief Description
---------|------------------
GetChart|Generates URL for image file of historical Zestimates for specified property
GetComps|Generates a list of comparable recent sales for a specified property
GetDeepComps|Same as above but with more variables for comp properties
GetSearchResults|Generates address, Zillow Property ID, and current Zestimate
GetDeepSearchResults|Same as above but more variables including lot size, year built, square footage, etc

## ZillowR Functions (cont'd)
Function | Brief Description
---------|------------------
GetMonthlyPayments|For a specific loan amount, returns estimated montly payment included and estimated principal and interest and tax if zip included
GetRateSummary|Returns current rates per loan type for specified state or national average
GetUpdatedPropertyDetails|Returns all of the home facts that have been edited by home's owner or agent
GetZestimate|Returns Zestimate for Zillow Property ID
zillow_web_service_id|Function to set your Zillow Web Service ID as global option

## ZillowR Functions 

Most of the functions return a named list with the following:

- **request** a list with the request parameters
- **message** a list of status codes and messages returned by the API
- **response** an XMLNode with the response from the API - no coercion is done to make it nice for R. This is why XML package is needed

## Getting Started
Always must start with loading your API Key, I am concealing mine but it's loaded in from a R script

```{r, echo=TRUE}
source("C:/Users/sazimmer/Documents/Voter/Programs/ZillowAPIKey.R")
# This 1-line R program contains the function pulling in my API key 
# I'm concealing this as it is a private key
# The program is simply:
# set_zillow_web_service_id('XXXXXXXXXX')
# where XXXXXXXXXX is my Zillow API key

```

## Addresses to play with

```{r, echo=TRUE, message=FALSE, results='asis'}
NCPolAdd <- read_csv("../Data/PoliticianAddresses.csv") %>%
  arrange(Org)
library(knitr)
kable(NCPolAdd[1:5,1:6])
```

## Addresses to play with (cont'd)

```{r, echo=TRUE, results='asis'}
kable(NCPolAdd[1:5,c(1,7:11)])
GovernorInfo <- NCPolAdd %>% filter(toupper(Org)=="GOVERNOR")
```

## Examples
```{r, echo=TRUE}
GDSROutput <- GetDeepSearchResults(address=GovernorInfo$StreetAddress, 
                     citystatezip = as.character(GovernorInfo$ZIP))
typeof(GDSROutput$request)
typeof(GDSROutput$message)
typeof(GDSROutput$response)

```

## Look at the request and message pieces
```{r, echo=TRUE}
GDSROutput$request
GDSROutput$message
```

## Look at the response piece
Let's pause and look at this in RStudio
```{r, echo=TRUE}
#Convert this to a list where we should
GovernorHomeList <- xmlToList(GDSROutput$response[["results"]][[1]])
names(GovernorHomeList)
GovernorHomeList$finishedSqFt
```

## What next?

- Rarely looking at just 1 address, need to learn the style of output
- Build your own custom functions to extract what you want - for us square footage
- Handle cases when query isn't found - not all addresses are on Zillow
- Use something like purrr to apply the same function to many addresses
